---
title: "DESeq2 Tutorial"
author: "Charlie George & Modified by Devika Agarwal"
date: "`r Sys.Date()`"
output: html_document
---

Useful resources 

- **Vignette** - http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html
- **Book** - http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

# install packages 

- If i didn't already have these installed in my environment - I would delete the # on lines 19-23 and 26-39 and run the code 

```{r}
# from cran 
# install.packages('tidyverse')
# install.packages("RColorBrewer")
# install.packages('pheatmap')
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

# from bioconductor
# BiocManager::install("ashr")
# BiocManager::install("apeglm")
# BiocManager::install("DESeq2")
# BiocManager::install("org.Mm.eg.db")
```

# load packages

- what do these packages do? Add comments to remind you what they do 
```{r}
library("pheatmap")   
library("RColorBrewer")
library(tidyverse)     
library(DESeq2)      
library(ashr)          
library(apeglm)      
library(org.Mm.eg.db)
```

# 1 Prepping our input data to make deseq2dataset object

## Import counts table

- We made this (or a very similar version) using featurecounts in week1 on the cluster
- raw data -> fastqc -> hisat2 -> featurecounts 

```{r}
# Import counts data to variable called featurecounts_table
# The file has a comment line we need to ignore at top of file 
featurecounts_table <- read_tsv("/t1-data/project/obds/shared/resources/3_r_stats_genomics/rnaseq_differential_expression/allsamples_counts.tsv.gz", comment = "#")


# lets save the gene info (Chr,Start,End,Strand, Length) columns as a seperate dataframe (df) incase we need it later
# call this table gene_info and set row names as Geneid
gene_info <- featurecounts_table %>%
                column_to_rownames("Geneid") %>%                     
                dplyr::select(Chr, Start, End, Strand, Length)


# create a counts table without the gene info (Chr,Start,End,Strand, Length) columns
# to get a clean counts table where:
# rows = genename, columns = samples
counts_table <- featurecounts_table %>%
                  column_to_rownames() %>%
                  dplyr::select() %>%
                  as.matrix()



# check the dimensions of this counts_table table using dim() -> should have 39,179 genes and 8 cols 


```

# Right now lets import and set up our table of sample information

```{r}
sample_info <- read_tsv("/t1-data/project/obds/shared/resources/3_r_stats_genomics/rnaseq_differential_expression/PRJEB18572_sample_info_subset.tsv")
sample_info

# What class is the cell_type currently? 



# We want to change it to a factor with 2 levels (the hypothesis we will test)
# Which should be our reference level? 


# what other columns might we want to include in our design that we want to change to factors? 



```

# Check the sample names are the same and in the same order in the sample_info and counts_table 

```{r}
# Column names of the count table and row names of the sample info table must be in the same order
# Check this 

# make sample_info row names from sample_id 

# remove .bam from colnames 


```

# Generate DESeqDataSet object

```{r}
# look up function to use from slides, find help and create an object called dds
dds <- 
  
```

## Investigate the dds object 

```{r}
# what does it look like? - e.g. use str(dds)


# how to access counts


# access colData i.e the sample info table


# access rowData


# access the design


# does it have any rowRanges()?


# what are its rownames and colnames?


# what attributes does it have?


```

## How to subset a DESeqDataSet Object 

```{r}
# practice subsetting dds object
# select first 3 columns


```

# Filter out unexpressed genes 

- It's good practice to filter out genes with 0 counts across all samples at the start 
- This reduces the size of the object and makes computations quicker without affecting the results as these genes are filtered out anyway in the analysis 
- Note that more strict filtering to increase power is automatically applied via independent filtering on the mean of normalized counts within the results function.


```{r}
# filter out genes that have 0 counts across all samples and save as an object called `dds_filtered`


# check dims of dds_filtered  
  
  
```

# Perform exploratory analysis on dds_filtered object 

1. transform dds_filtered object using `rlog()` functions
    - name your new variables `rlog_transformed_dds` 
    - we want to set `blind = TRUE` as because we are doing a QC step we do not want the function to take into account the experimental design, we want it to be an unbiased look at how the samples cluster
    - check the class of the rlog_transformed_dds objects 
    - you can look at the tranformed counts table using `assay()` 

NOTE we only use this tranformed data for QC not for downstream analysis or differential expression itself!!


```{r}
#transform data using `rlog()` functions
rlog_transformed_dds <- 


# Notes on `blind` parameter: 
  # blind = TRUE is default
  # blind = TRUE is not the appropriate choice if you expect many or majority of genes to have large differences in counts explainable by   experimental design
  # Mike Love recommends blind = FALSE for visualisation, clustering and machine learning but blind = TRUE for QC
  # Experimental design is only used for calculating the gene-wise dispersion estimates in order to fit a trend line through the dispersions over the mean

  
#check the class of these variable objects 

  
#get and look at the tranformed counts table using `assay()`


```

# Optional: If you have time do the tranformation again but using the vst() function 

- As a guide 
    - rlog - more robust than VST when size factors vary widely (10x)
    - VST - less sensitive to high count outliers than rlog, quicker - recommended most often 

```{r}
# get vst transformed data using vst() 
vst_transformed_data <- 

```


# Now plot heatmap of sample-sample correlations using rlog transformed count data and cor()


```{r}
# get correlation matrix of the rlog_transformed counts matrix to get correlations between samples 


# use pheatmap to plot the correlation matrix 

```

# try ploting a heatmap of sample distances 

```{r}
# get sample distances -> Annotated what this code is doing?
sampleDists <- dist(t(assay(rlog_transformed_dds))) 
sample_distances_matrix <- as.matrix(sampleDists) 

pretty_blue_colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

pheatmap(sample_distances_matrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists, 
         color = pretty_blue_colors)
```

What do you notice here!? what does it tell us about our data? 

Answer: 



## You can subset a dds object and interact with it like you would a DataFrame

```{r}
# remove the two suspect columns using [,] notation - remove suspicious samples 
rlog_transformed_dds_subset <- rlog_transformed_dds[,1:6]

# lets replot the correlation heatmap with the subsetted data - copy code and adjust from above


```

What do you notice now about the data and the clustering?


# Now lets use the plotPCA function from DESeq2 to plot PCA of rlog_transformed_data_subset

```{r, fig.asp=1}
# plot pca of rlog_transformed_dds using plotPCA()


# what does the intgroup parameter do? have a go at changing it...


# how many variable genes is it using for pca? 


# plot the pca for all genes 

```


What does it tell us about our sample?

Answer: 


------------- This is the end of our Exploratory data analysis -----------------

We now go back to our original dds object with the raw counts and can update it with anything we need to from the exploratory analysis 
    e.g. - update design with variables in sample info table we want to regress out 
         - remove samples (poor quaility or those with name switches etc)
         
# Revisiting our inital data 

What do we need to do to our data??? 

## Update dds_filtered object if needed - call it dds_filtered_updated

```{r}

# do we need to remove/edit samples? 

# do we want to update the design? Is this nessecary in this case?
# e.g. design(dds) <- ~replicate + condition

```


# Now we can start our actual DE!!!! yay 

Run deseq2 modelling using one function (DESeq())

```{r}

# look at dds slots then run deseq2 function 
# what has happened to your dds slots?

# run deseq2 modelling on the updated dds object in one function using DEseq() 
dds_filtered_updated <- DESeq(dds_filtered_updated)

# then extract the results
results_obj <- results(dds_filtered_updated)

# sumarise results 
summary(results_obj)

```

Have a look at the outputs 

# Now we will run the analysis step by step to see whats going on 

## 1. Estimate size factors using estimateSizeFactors()

```{r}
# Calculate the size factors for each sample using estimateSizeFactors()

# Access the size factors from the dds_filtered_updated object

# Do you see any systematic differences in size factors between groups?

```

# Optional: Plot size factors as barchart using ggplot
```{r}
```

## Get counts table normalised using size factors

```{r}
# get counts table normalised using size factors - call it normalized_counts


# check the class of the output

```

# 3. Estimate dispersions for each gene

```{r}
# Obtain dispersion estimates for each gene using estimateDispersions() 

# Plot the per-gene dispersion estimates (DESeq2 has a helper function for this)

```


## 4. Perform Wald test

```{r}
# Tests for significance of coefficients in a negative binomial GLM, using previously calculated size factors and dispersion estimates and WaldTest

```


# So we have the actual steps that we've gone through to get to DE in one place generate the dds object again and use the DESeq() function to perform steps 1-3 in one go

```{r}
# generate dds object 

# filter rows with 0 counts 

# remove unwanted samples 

# make DESeqDataSet (dds) object 

# run DESeq() 

```

# Get the results from your dds object - call it `results_obj`

```{r}
# What do the results() and summary() do, what arguments can you specify? 


```

# Optional - if we have time: Access the coefficients of the Negative Binomnial GLM – see vingette("DESeq2")
```{r}

```

# Plot normalised counts for ENSMUSG00000041779
 
 checkout the 'plotCounts()` function
 
```{r}
# Plot normalised counts for gene ENSMUSG00000041779

```

# Access the results table for the comparison between CD8 and CD4 T cells . Youu can also try to CD4 vs CD8 (i.e. switch the reference) - you could also try this for replicate 1 vs replicate 3 as a control 

checkout `resultsNames()`  & `results()`

```{r}

```

Table should contain log2 fold changes and p-values for all genes

# Plot a histogram of the raw and BH-adjusted p-values using ggplot – do they look as expected? for CD8 vs CD4

```{r}

```


# Generate an MA plot of the log2 FC values for all genes using DESeq2 plotMA()

```{r}


```

# Shrink the log2 FC values using the normal, apeglm and ashr methods -> see the vignette – which shall we choose? 

```{r}

# call your results table  called shrunkenLFC_dds_results_obj

```


# Generate a results table (one shrinkage method) containing gene symbols and (if time) entrez ids 

```{r}

# convert shrunkenLFC_dds_results_obj to a dataframe and call it shrunkenLFC_dds_results_df

# Use the org.Mm.eg.db package to get gene symbols for the ensembl gene ids (hint:you did this on bioconductor day) 

# How many Ensembl IDs are not assigned an gene symbol and how many gene symbols are duplicated?



keytypes(org.Mm.eg.db)      #to remind us of what keytypes are in this particular db

# use mapIds() to get 1:1 mapping for SYMBOL, ENTREZ and MGI

# add mapId vectors to  shrunkenLFC_dds_results_df
```


# Export all the results to CSV

```{r}
write_csv(shrunkenLFC_dds_results_df, "CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")
```

# Filter res_df and save as CSV

```{r}
#Filter the results table for padj < 0.05 and absolute log2 FC > 1, and write to a CSV file
res_0.05 <- shrunkenLFC_dds_results_df %>% 
    dplyr::filter(padj < 0.05 & abs(log2FoldChange) > 1)

#Get upregulated genes - Filter the results table for padj < 0.05 and log2FoldChange FC > 1, and write to a CSV file
res_0.05_up <- shrunkenLFC_dds_results_df %>% 
    dplyr::filter(padj < 0.05 & log2FoldChange > 1)

#Get doownregulated genes - Filter the results table for padj < 0.05 and log2FoldChange FC < 1, and write to a CSV file
res_0.05_down <- shrunkenLFC_dds_results_df %>% 
    dplyr::filter(padj < 0.05 & log2FoldChange < -1)

write_csv(res_0.05, 
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1.csv")

write_csv(res_0.05_up,
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")

write_csv(res_0.05_down,
          "CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")
```


Well done! That is the basics of exploratory data analysis for bulk RNAseq and differential gene expression - these statistical methods underly the foundations for single cell differential expression analysis, although scRNA-seq has a few extra paramterisations and complications to take into account. 

We will now progress to pathway/geneset analysis to get some biological meaning from our DE gene list....

