---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer:

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
diamonds
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut))
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut)) +
geom_smooth()
  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut)) +
geom_smooth()

```

- Assign that last plot to an object called `obds_diamonds`.

```{r}

obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) +
geom_point(size = 0.7, aes(colour = cut)) +
geom_smooth()

```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:
```{r}
head(ChickWeight)
```

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ChickWeight
ggplot(ChickWeight, aes(x = Time, y = weight)) +
geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
geom_point(aes(colour = Diet))

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
geom_point() +
geom_smooth(method = lm)


```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
geom_point() +
geom_smooth(method = lm) +
  facet_wrap(~Diet)


```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
geom_point() +
geom_smooth(method = lm) +
  facet_wrap(~Diet)


```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1))



```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar() +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1),
    axis.title=element_text(size = rel(2))
    )+
labs(x = "Order", y = "Observations")



```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus)) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1),
    axis.title=element_text(size = rel(2)),
    legend.key.size = unit(1.2, "mm")
    )+
labs(x = "Order", y = "Observations")




```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}








```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus)) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1),
    axis.title=element_text(size = rel(2)),
    legend.key.size = unit(1.2, "mm"),
    legend.text = element_text (size = rel(0.5))
    )+
labs(x = "Order", y = "Observations") +
guides(fill = guide_legend(ncol = 3))







```

- Add a contour of thin black lines to the bars.

```{r}

ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus), colour = "black", size = 0.1) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1),
    axis.title=element_text(size = rel(2)),
    legend.key.size = unit(1.2, "mm"),
    legend.text = element_text (size = rel(0.5))
    )+
labs(x = "Order", y = "Observations") +
guides(fill = guide_legend(ncol = 3))









```

- Assign that last plot to an object called `obds_msleep`.

```{r}


obds_msleep <- ggplot(msleep, aes(x = order)) +
  geom_bar(aes(fill = genus), colour = "black", size = 0.1) +
  theme(
    axis.text.x = element_text(angle = 45, size = rel(0.9), hjust = 1, vjust = 1),
    axis.title=element_text(size = rel(2)),
    legend.key.size = unit(1.2, "mm"),
    legend.text = element_text (size = rel(0.5))
    )+
labs(x = "Order", y = "Observations") +
guides(fill = guide_legend(ncol = 3))







```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=5, fig.width=7}
library(cowplot)

#cowplot::plot_grid(obds_diamonds, obds_chickweight, obds_msleep, labels = c("A", "B", "C"), ncol = 2, nrow =2)

#plot_grid(plot1, plot2, plot)

first_row <- plot_grid(obds_diamonds, obds_chickweight, labels = c("A", "B"), ncol = 2, nrow =1)
second_row <- cowplot::plot_grid(first_row, obds_msleep, labels = c("", "C"), nrow = 2)

second_row
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)

obds_diamonds + obds_chickweight + obds_msleep

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)





```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave(   
  "superplot.pdf",
  second_row,
  width = 10,
  height = 10
  )
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```

