---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- read10xCounts(c(PBMC5k = "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix"))
  
  

sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
sce <- sce[,colSums(counts(sce)) !=0] #cannot use colSums here: select the counts matrix and choose only non-zero values. Could have selected >0.
sce
```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(counts(sce), lower = 100) #could have used ignore = 0 to replace previous code (but this is a worse option...)
out
```

**How do you read the output? Which droplets would you consider empty?**

> Answer: One row for everything including the barcodes with <100 UMI that weren't tested (hence lots of Nas). TOtal tells us the total UMI for that barcode.
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
is.cell <- out$FDR <= 0.001 #false discovery rate, 
sum(is.cell, na.rm=TRUE)
#could have done a historgram to pick out
sce_filtered_emptydrops <-   sce[,which(is.cell), drop=F] #subsetting the matrix to the cell containing droplets using 'which()' to handle NAs smoothly
sce_filtered_emptydrops #1202 columns
table(Sig=is.cell, Limited=out$Limited)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
library(scuttle)

sce <- scuttle::addPerCellQC(sce)
sce$is.cell <- is.cell
ncol(sce) #234600 columns
length(is.cell) # 234600

colData(sce)


ggplot(as.data.frame(colData(sce))   ) +
  geom_violin(aes(x = is.cell, y = sum))+
  labs(x = "empty droplet", y = "library size")+
  coord_cartesian(ylim=c(0,20000))+
  theme_classic()+
  geom_violin(aes(x = is.cell, y = sum), scale = "width")+
    geom_jitter(aes(x = is.cell, y = sum))
  
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- read10xCounts(c(PBMC_filtered = "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/filtered_feature_bc_matrix/"))

sce_filtered_cellranger

colData(sce_filtered_cellranger)

all_barcodes <- sce$Barcode

length(all_barcodes)

compare_table <- tibble(
  emptydrops = all_barcodes %in%  sce_filtered_emptydrops$Barcode  ,
  cellranger = all_barcodes %in%  sce_filtered_cellranger$Barcode
)

table(compare_table) #This table tells us which barcodes were called as empty in cellranger vs in our emptyDrops vs called as good cells. Allows identification of disagreement between the two methods.
compare_table
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_cellranger)

sce_scdblfinder

colData(sce_scdblfinder)
#the higher the score, the more likely they are doublets

table(sce_scdblfinder$scDblFinder.class)

hist(sce_scdblfinder$scDblFinder.score, breaks = 100)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
sce_filtered_scdblfinder <- sce_scdblfinder[, sce_scdblfinder$scDblFinder.class=="singlet"]

sce_filtered_scdblfinder

table(sce_filtered_scdblfinder$scDblFinder.class) 
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder)



plot_data <- ggplot(as.data.frame(colData(sce_scdblfinder)))+
  geom_violin(aes(x=scDblFinder.class, y = sum), scale = "width") + labs(x="Singlet or Doublet", y = "Library Size")

plot_data

colData(sce_scdblfinder)
```

```{r}
is.mt <- grep("^MT-", rowData(sce_scdblfinder)$Symbol)

is.mt

sce_scdblfinder <- scuttle::addPerCellQC(sce_scdblfinder, subsets = list("is.mt" = is.mt))

colData(sce_scdblfinder)

mt.plot <- ggplot(as.data.frame(colData(sce_scdblfinder))) +
                    geom_violin(aes(x = scDblFinder.class, y = subsets_is.mt_percent)) +
                    labs(x = "Singlet or Doublet", y = "Percentage mitochondrial")

mt.plot




  
```

```{r}
sce_scdblfinder <- 
sce_scdblfinder <- 
plot_data <- tibble(
  
  
)

ggplot(   ) +
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_scdblfinder,
                              z=NULL,
                              batch=NULL,
                              maxIter=500
                              )
  
  
colData(sce_decontX)
str(metadata(sce_decontX))
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
dim(sce_decontX)
class(sce_decontX)

reducedDimNames(sce_decontX) #"decontX_UMAP"
umap <- reducedDim(sce_decontX, "decontX_UMAP")

umap

plotDimReduceCluster(x = sce_decontX$decontX_clusters,
                     dim1 = umap[,1],
                     dim2=umap[,2]
                     )

#give as vector
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX,
                         batch = NULL,
                         colorScale = c("blue", "green", "yellow", "orange", "red"),
                                        size = 1)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}
sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts")
rownames(sce_decontX) <- uniquifyFeatureNames(
  ID = rownames(sce_decontX),
  names = rowData(sce_decontX)[["Symbol"]]
)

sce_decontX

#take gene symbol, use that as row name, otherwise keep ensembl ID, if duplicated, uniquefy it with geneID. So rowname  

assayNames(sce_decontX)

listofgene<- c("CD3D", "CD3E", "GNLY", "LYZ", "S100A9", "CD79A", "CD79B", "MS4A1")
listofgene

plotDimReduceFeature(x = assay(sce_decontX, "logcounts"),
                     features = listofgene,
                     dim1=umap[,1],
                     dim2=umap[,2])
  

  
  
  
  
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list(
  Tcell_Markers = c("CD3E", "CD3D"),
  Bcell_Markers = c("CD79A", "CD79B", "MS4A1"),
  Monocyte_Markers = c("S100A8", "S100A9", "LYZ"),
  NKcell_Markers = "GNLY"
)
cellTypeMappings <- list(
  Bcells = 4,
  Monocytes = 2,
  Tcells = c(1, 3),
  NKcells = c(5, 6)
)
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = "counts"
)
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r}
plotDecontXMarkerExpression(
  
  
  
  
)
```

